### 2018-11-07 Sabanci_talk/Sabanci_talk.tex
Improved many slides, added new figs

\begin{frame}{Outline}
\begin{itemize}
 \item Motivation
 \item Gravitational waves
 \item Binary neutron star inspirals
 \item Gravitational-wave detectors
 \item Event rates 
\end{itemize}
\todo{finish this slide remove later if short on time}

@@

\begin{frame}{Motivation: Multi-messenger Astronomy}
\begin{itemize}
 \item \emph{Multi}: two fundamentally different waves from common events.
 \item[]\quad Gravitational waves (GWs): ``hearing''  the universe.
 \item[]\quad Electromagnetic waves (EWs): ``seeing'' the universe.
 \item \red{GW170817}-\bl{GRB170817A}-\textcolor{brown}{AT2017gfo}: the ``event'' \\
 \quad inspiral + merger of two neutron stars (GWs) \\
 \ \ = short-hard gamma-ray burst + kilonova (EWs)
 \item[] Broad-band in gravitational waves: 10Hz - 2000Hz
 \item[] Ultra broad-band in electromagnetic waves: gamma rays to radio.
 \item A simple question \\
{\small ``What can we offer to the astronomy community with future GW detections?''}
 \end{itemize}

@@

\begin{frame}{Gravitational waves: Einstein 1916}
  \begin{center}
\begin{tikzpicture}
            \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[height=3.5cm]{figs/Einstein1916.png}};
            %\draw<1>[red,ultra thick,rounded corners] (1.6,1) rectangle (\textheight-1cm,5);
           \uncover<3->{\draw<3->[red,thick] (1.7,0.04) rectangle (3.1,0.27);}
        \end{tikzpicture}
        \\
         \uncover<2->{Perturbations of spacetime with speed $=c$, sourced by accelerating masses.}
         \end{center}
          \begin{itemize}
 \item  \uncover<3->{Spacetime metric $g_{\mu\nu} = \eta_{\mu\nu} + h_{\mu\nu}$, \quad with $| h_{\mu\nu} | \ll 1$.}
 \item  \uncover<4->{Insert into $G_{\mu\nu} = 0$. Keep only $\mathcal{O}(h)$. Pick a gauge.} %(e.g. Lorenz).}
  \uncover<5->{\[ \Rightarrow\quad \Box \bar{h}_{\mu\nu}= 0, \quad \text{wave equation!} \]}
  \item \uncover<6->{\alert{Plane-waves:} $ \bar{h}_{\mu\nu} = \Re\left[A_{\mu\nu}\, \text{e}^{i k_\mu x^\mu}\right] =\Re\left[A_{\mu\nu}\, \text{e}^{-i\omega (t-z/c )}\right]$.}
 \end{itemize}
\end{frame}

\begin{frame}{Gravitational waves: d.o.f.'s}
$A_{\mu\nu} $ is the polarization tensor.
\begin{itemize}
 \uncover<2->{\item $\{ \bar{h}_{\mu\nu}, A_{\mu\nu} \}$: $4\times4$, symmetric \uncover<3->{$\hspace{1.5cm}\Rightarrow \f{4\times 5}{2}\ \ \ = 10$ d.o.f.}}
 \uncover<4->{\item Gauge condition: $\nabla_\mu \bar{h}^{\mu\nu} =  A_{\mu\nu} k^\nu= 0$ \uncover<5->{$\Rightarrow10-4 = \, 6$ d.o.f.}}
 \uncover<6->{\item Residual gauge freedom \hspace{2.25cm} \uncover<7->{$\Rightarrow 6-4 \ \ = \alert{2}$ \alert{d.o.f.}}}
 %\uncover<7->{\vspace{2mm}\begin{center}$ \bar{h}'_{\mu\nu} = \bar{h}_{\mu\nu} - \xi_{\mu,\nu} - \xi_{\nu,\mu} + \eta_{\mu\nu} \nabla_\alpha \xi^\alpha . $ \end{center}\vspace{2mm}}
\end{itemize}
\uncover<7->{\quad \ Consistent with $\pm 2$ helicities of a \bl{massless} spin 2 boson.\\}
\uncover<8->{$\Rightarrow$ \alert{2} POLARIZATIONS (\alert{transverse}): \uncover<10->{$\begin{array}{l}A_{xx}=-A_{yy} = \red{h_+}  \\ A_{xy}=\ \ A_{yx} = \red{h_\times}  \\ \end{array}$}}
\\
\todo{Replace with Maggiore's discussion?}
{\begin{center} \includegraphics[height=2.79875cm]{figs/white_box.png}\end{center}}
\end{frame}

\begin{frame}{Gravitational waves: d.o.f.'s}
%\begin{center}
%\animategraphics[loop,controls,width=3cm]{12}{GWs_frame-}{0}{29} 
%\end{center}
$A_{\mu\nu} $ is the polarization tensor.
\begin{itemize}
{\item $\{ \bar{h}_{\mu\nu}, A_{\mu\nu} \}$: $4\times4$, symmetric {$\hspace{1.5cm}\Rightarrow \f{4\times 5}{2}\ \ \ = 10$ d.o.f.}}
 {\item Gauge condition: $\nabla_\mu \bar{h}^{\mu\nu} =  A_{\mu\nu} k^\nu= 0${ $\Rightarrow10-4 = \,6$ d.o.f.}}
 {\item Residual gauge freedom \hspace{2.25cm}{ $\Rightarrow 6-4 \ \ = \alert{2}$ \alert{d.o.f.}}}
 %{\vspace{2mm}\begin{center}$ \bar{h}'_{\mu\nu} = \bar{h}_{\mu\nu} - \xi_{\mu,\nu} - \xi_{\nu,\mu} + \eta_{\mu\nu} \nabla_\alpha \xi^\alpha . $ \end{center}\vspace{2mm}}
\end{itemize}
{\quad \ Consistent with $\pm 2$ helicities of a \bl{massless} spin 2 boson.\\}
{$\Rightarrow$ \alert{2} POLARIZATIONS (\alert{transverse}): {$\begin{array}{l}A_{xx}=-A_{yy} = \red{h_+}  \\ A_{xy}=\ \ A_{yx} = \red{h_\times}  \\ \end{array}$}}
\\
\todo{Replace with Maggiore's discussion?}
{\begin{center} \animategraphics[loop,controls,height=2.4cm]{12}{GWs_plus-}{0}{20} \hspace{1cm} \animategraphics[loop,controls,height=2.4cm]{12}{GWs_cross-}{0}{20}\end{center}}
\end{frame}

\begin{frame}{Making gravitational waves}{non-spherical, accelerated motion}
 \alert{Inspirals}, GRBs, bumps on neutron stars, supernovae, the Big Bang. \\% (BICEP 2).\\
 \uncover<2->{Gravitational-wave luminosity: \uncover<3->{\alert{quadrupole} radiation}  (leading-order)}
 \uncover<3->{\begin{center} $ \boxed{\dot{E}=L_\text{GW} = \f{G}{c^5} \langle (\partial_t^3 \bar{Q}_{ij})^2 \rangle}$\\\vspace{2mm}}
 \uncover<4->{\begin{small}$ \quad \bar{Q}_{ij} = Q_{ij}-\frac{1}{3}\delta_{ij}Q^k_k, \quad Q^{ij} = \int \rho(t,{\mathbf x})x^i x^j d^3 x $ \end{small} \end{center}}
 \uncover<5->{{\bf Binary systems:} 2 points masses in circular orbit. }

\begin{frame}{Making gravitational waves}{non-spherical, accelerated motion}
 \alert{Inspirals}, GRBs, bumps on neutron stars, supernovae, the Big Bang. \\%  (BICEP 2).\\
 Gravitational-wave luminosity: \alert{quadrupole} radiation (leading-order)
 \begin{center} $ \boxed{\dot{E}=L_\text{GW} = \f{G}{c^5} \langle (\partial_t^3 \bar{Q}_{ij})^2 \rangle}$\\ \vspace{2mm}
 \begin{small}$ \quad \bar{Q}_{ij} = Q_{ij}-\frac{1}{3}\delta_{ij}Q^k_k, \quad Q^{ij} = \int \rho(t,{\mathbf x})x^i x^j d^3 x $ \end{small} \end{center}
 {\bf Binary systems:} 2 points masses in circular orbit.\todo{$R\to r$ in vid}
 \begin{tikzpicture}[overlay,remember picture]
\onslide<5->\node (img0 )[anchor=center,scale=1,opacity=1] at ([shift={(-3.8cm,-2.8cm)}]current page.center) {\animategraphics[loop,controls,height=2.4cm]{12}{Circ_Orb-}{0}{20}};
\onslide<2->\node (t1 )[anchor=center,scale=1,opacity=1] at ([shift={(2cm,-1.3cm)}]current page.center) {$\mu=m_1 m_2/M^2,\ x(t) = r \cos\Omega t, \ y(t) = r \sin\Omega t $};
\onslide<3->\node (t2 )[anchor=center,scale=1,opacity=1] at ([shift={(2.25cm,-2.2cm)}]current page.center) {$ Q^{ij} =  2 M\, x^i x^j, \qquad \boxed{\dot{E} = \f{32}{5}\f{G\mu^2}{c^5}r^4 \Omega^6}$};
\onslide<4->\node (t3 )[anchor=center,scale=1,opacity=1] at ([shift={(2.5cm,-3.2cm)}]current page.center) {Enhancement due to \bl{$e$} (Peters \& Mathews).};
\onslide<5->\node (t3 )[anchor=center,scale=1,opacity=1] at ([shift={(2cm,-4cm)}]current page.center) {$ \bl{\f{\left(1 + \f{73}{24}e^2+ \f{37}{96} e^4 \right)}{(1-e^2)^{7/2}}}$};
\end{tikzpicture}
\end{frame}

 \begin{frame}{Compact Binary Inspirals}
 Notation:
 \begin{itemize}
  \item $\red{f} $: GW frequency, \hspace{1.5cm} $\Longrightarrow\quad \omega\equiv2\pi f = 2\Omega$.
  %\item $\omega\equiv2\pi f = 2\Omega$: GW angular frequency.
  \item $t$: observer/detector time \quad$ \Longrightarrow\quad\dot{Y}= dY/dt$.
 \end{itemize}

  Assumptions:
  \begin{itemize}
   \item Quasi-circular orbits: $e \ll 1$ and $\tfrac{|\dot{r}|}{r\Omega} <10^{-3} $ at $f=100\,$Hz.
   \item Quadrupolar GWs only: higher modes suppressed by $\tfrac{m_1-m_2}{M}\tfrac{v^2}{c^2}$
   \item Use Newtonian/post-Newtonian theory ($\le$3.5{\bf PN}).
   \item Terminate inspiral at Schwarzschild ISCO \todo{add ISCO curve?}
   \begin{footnotesize}
   \[f_\text{ISCO} = \f{c^3}{6^{3/2}\pi G M} \simeq 1571\,\text{Hz}\, \left(\f{2.8M_\odot}{M}\right)\] 
   \end{footnotesize}
   \end{itemize}
   %
   For binary neutron stars ({\bf BNS})
   \begin{itemize}
   \item Point masses: tidal effects for $f \gtrsim 100\,$Hz
   \item Set $m_1 =m_2 = 1.4 M_\odot$, constraints: $1.2 M_\odot \lesssim m_\text{NS}\lesssim 1.6M_\odot$.
   \item Neglect spins: spin-orbit $\tfrac{\dot{E}_\text{SO}}{\dot{E}} \sim \tfrac{v_s}{c}\tfrac{v}{c}\tfrac{R_\text{NS}}{r}$, spin-spin $\sim \tfrac{v^4}{c^4}$
   \end{itemize}

@@

\begin{frame}{Compact Binary Inspirals}{Inspiral time \todo{Add $r(t)$ plot? }}
Integrate $\dot{f} \sim f^{11/3} \quad \Longrightarrow \quad \int dt \sim \int f^{-11/3}df \quad\Rightarrow\quad t \sim \red{f^{-8/3}}$
\\
\vspace{1mm}
Fix integration constant $\tau = t_\text{coal} - t >0$. \\
\vspace{2mm}
Introduce \red{inspiral time}, i.e, time to BNS \bl{coal}escence
%
\begin{align}
 \red{\tau_\text{insp}(f)} %&= \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,{f^{-8/3}}\nn\\
 &\simeq \bl{16.72\,\text{minutes}} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,{\left(\f{10\,\text{Hz}}{\red{f}}\right)^{\red{8/3}}\quad \red{(2)}}\nn
\end{align}
Number of GW cycles to coalescence: $\int \tfrac{f}{\dot{f}}df\sim \red{f^{-5/3}}$ %$\int df {f}/\dot{f}$
\begin{align}
 \red{\mathcal{N}_\text{cyc}(f)} %&=\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right]\nn \\
 &\approx 1.605\times \bl{10^4}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}{ \left(\f{10\,\text{Hz}}{\red{f}} \right)^{\red{5/3}}}\quad \red{(3)}\nn
\end{align}
%
\begin{tikzpicture}[overlay,remember picture]
\node[draw, red,thick,minimum width=8.2cm,minimum height=1.3cm] (b) at (7.,3.35){};
\node[draw, red,thick,minimum width=10cm,minimum height=1.3cm] (b) at (6.2,1){};
\end{tikzpicture}
%
\[ 
\tfrac{\dot{r}}{r}\sim \tfrac{\dot{f}}{f} \Longrightarrow \red{r(\tau)}=r_i \left(\f{\red{\tau}}{\tau_i}\right)^{\red{1/4}}
\]

@@

\begin{frame}{Gravitational waves from BNS}
TT gauge, multipole, slow-velocity expansions give \todo{add suppl. slide}
\begin{small}
\[
 h_{ij}^\text{TT} = \tfrac{2}{D}\tfrac{G}{c^4} \ddot{Q}_{ij}^\text{TT}%(t-z/c)
\]
%
 \begin{align*}
 h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi(t)],\\
 h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi(t)],
\end{align*}
\end{small}
%
%
$\iota =\,$ orbital \bl{inclination}, and
\vspace{-3mm}
%
\begin{small}
\begin{align*}
h_c(t) %&=\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3}\nn\\
&\approx 1.9\times \bl{10^{-23}} \left(\f{40\,\text{Mpc}}{D}\right)\left(\f{M_c}{1.219 M_\odot}\right)^{5/3}\left(\frac{\red{f}}{10\,\text{Hz}}\right)^{\red{2/3}} \nn\\
\Phi(t) &=\int_{t_i}^t dt' \omega(t') + \text{PN corrections}
\end{align*}
\end{small}\\
%
Frequency increases as $f^{11/3}$\\
Amplitude increases as $f^{10/3} \Longrightarrow$ \red{Chirp!} \todo{Insert chirp movie}

 \begin{frame}{GW Detectors: \red{Interferometers}}
  Two geodesics separated by $\xi^\mu$:
  %\[
  % \frac{D^2 \xi^\alpha}{D\tau^2} = -R^\alpha_{\beta \gamma\delta}\, \xi^\gamma u^\beta u^\delta 
  %\]
$ \tfrac{d^2 \xi^i}{d\tau^2} = -R^i_{0j0} \xi^j\left(\tfrac{dx^0}{d\tau}\right)^2$ \hfill (detector frame)\\
%\hspace{4cm} $\Longrightarrow \ddot{\xi^i} = -c^2 R^i_{0j0} \xi^j $ \hfill (detector time)\ \\
\vspace{-3mm}
\[
\hspace{4.5cm} \boxed{\ddot{\xi^i} = \frac{1}{2} \ddot{h}_{ij}^\text{TT} \xi^j} \hspace{2.5cm} \text{(detector time)}
\]
%
%
GW along $\mathbf{z}$: 
\[
 h_{ij}^\text{TT} = \left(\begin{array}{cc}
                           h_+ & h_\times \\
                           h_\times & - h_+
                          \end{array}\right) \cos\left(\omega\left( t-\tfrac{z}{c}\right)\right)
\]
Consider mirrors at $(0,0)_1, (L,0)_2$ and $(L,L)_3$
\begin{small}
\begin{align*}
  s_{12} &= \int_1^2 \tfrac{ds}{dx}dx = (x_2-x_1) \left[ 1 + h_+ \cos(\omega t)\right]^{1/2}\hspace{4cm} \\
			      & \approx L \left[ 1+\tfrac{1}{2} h_+ \cos\omega t \right] \\
s_{13} &\approx L \left[ 1-\tfrac{1}{2} h_+ \cos\omega t \right]
\end{align*}
\end{small}
%
\begin{tikzpicture}[overlay,remember picture]
%\node (img0 )[anchor=center,scale=1,opacity=1] at ([shift={(3.8cm,-2.8cm)}]current page.center) {\animategraphics[loop,controls,height=2.39cm]{12}{LIGO_int-}{0}{102}};
\end{tikzpicture}
 \end{frame}

@@

\begin{frame}{Optimal Signal-to-Noise Ratio}
 RMS Average over the angles \\
 Then introduce SNR for both LIGO and ET
\end{frame}

>>>>

\begin{frame}{Outline}
\begin{itemize}
 \item Motivation
 \item Gamma-ray bursts
 \item Gravitational waves
 \item Binary neutron star inspirals
 \item Gravitational-wave detectors
 \item Event rates 
\end{itemize}
\todo{finish this slide remove later if short on time}

@@

\begin{frame}{Motivation: Multi-messenger Astronomy}
  \begin{itemize}
    \item \emph{Multi}: two fundamentally different waves from common events.
    \item[]\quad Gravitational waves (GWs): ``hearing''  the universe.
    \item[]\quad Electromagnetic waves (EWs): ``seeing'' the universe.
    \item \red{GW170817}-\bl{GRB170817A}-\textcolor{brown}{AT2017gfo}: the ``event'' \\
    \quad inspiral + merger of two neutron stars (GWs) \\
    \ \ = short-hard gamma-ray burst + kilonova (EWs)
    \item[] Broad-band in gravitational waves: 10Hz - 2000Hz
    \item[] Ultra broad-band in electromagnetic waves: gamma rays to radio.
    \item A simple question \\
    {\small ``What can we offer to the astronomy community with future GW detections?''}
  \end{itemize}
\end{frame}

\begin{frame}{Gamma-ray bursts}{Most energetic explosions in the universe}
 \todo{Use Pawan's slides}\\
 Brief history: spy satellites ...\\
 Two main types, short long,  etc.

@@

\begin{frame}{Neutron Stars}{Most extreme stars in the universe}
  Stellar evolution -> white dwarf, NS -> BH \\
  Density, mass, radius, spin, nuclear equation of state\\
  Tea spoon of neutron star matter trillion tons
\end{frame}

\begin{frame}{Gravitational waves: Einstein 1916}
  \begin{center}
\begin{tikzpicture}
            \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[height=3.5cm]{figs/Einstein1916.png}};
            %\draw<1>[red,ultra thick,rounded corners] (1.6,1) rectangle (\textheight-1cm,5);
           \uncover<3->{\draw<3->[red,thick] (1.7,0.04) rectangle (3.1,0.27);}
        \end{tikzpicture}
        \\
         \uncover<2->{Perturbations of spacetime with speed $=c$, sourced by accelerating masses.}
         \end{center}
          \begin{itemize}
 \item  \uncover<3->{Spacetime metric $g_{\mu\nu} = \eta_{\mu\nu} + h_{\mu\nu}$, \quad with $| h_{\mu\nu} | \ll 1$.}
 \item  \uncover<4->{Insert into $G_{\mu\nu} = 0$. Keep only $\mathcal{O}(h)$. Pick a gauge.} %(e.g. Lorenz).}
  \uncover<5->{\[ \Rightarrow\quad \boxed{\Box \bar{h}_{\mu\nu}= 0}, \quad \text{wave equation!} \]}
  \item \uncover<6->{\alert{Plane-waves:} $ \bar{h}_{\mu\nu} = \Re\left[A_{\mu\nu}\, \text{e}^{i k_\mu x^\mu}\right] =\Re\left[A_{\mu\nu}\, \text{e}^{-i\omega (t-z/c )}\right]$.}
 \end{itemize}
\end{frame}

\begin{frame}{Gravitational waves: d.o.f.'s}
$A_{\mu\nu} $ is the polarization tensor.
\begin{itemize}
 \uncover<2->{\item $\{ \bar{h}_{\mu\nu}, A_{\mu\nu} \}$: $4\times4$, symmetric \uncover<3->{$\hspace{1.5cm}\Rightarrow \f{4\times 5}{2}\ \ \ = 10$ d.o.f.}}
 \uncover<4->{\item Gauge condition: $\nabla_\mu \bar{h}^{\mu\nu} =  A_{\mu\nu} k^\nu= 0$ \uncover<5->{$\Rightarrow10-4 = \, 6$ d.o.f.}}
 \uncover<6->{\item Residual gauge freedom \hspace{2.25cm} \uncover<7->{$\Rightarrow 6-4 \ \ = \alert{2}$ \alert{d.o.f.}}}
 %\uncover<7->{\vspace{2mm}\begin{center}$ \bar{h}'_{\mu\nu} = \bar{h}_{\mu\nu} - \xi_{\mu,\nu} - \xi_{\nu,\mu} + \eta_{\mu\nu} \nabla_\alpha \xi^\alpha . $ \end{center}\vspace{2mm}}
\end{itemize}
\uncover<7->{\quad \ Consistent with $\pm 2$ helicities of a \bl{massless} spin 2 boson.\\}
\uncover<8->{$\Rightarrow$ \alert{2} POLARIZATIONS (\alert{transverse}): \uncover<10->{$\begin{array}{l}A_{xx}=-A_{yy} = \red{h_+}  \\ A_{xy}=\ \ A_{yx} = \red{h_\times}  \\ \end{array}$}}
\\
\begin{tikzpicture}[overlay,remember picture]
\node (img0 )[anchor=center,scale=1,opacity=1] at ([shift={(-3.5cm,-2.5cm)}]current page.center) {\animategraphics[loop,controls,height=2.6cm]{12}{GWs_plus-}{0}{20}};
\node (img1 )[anchor=center,scale=1.,opacity=1] at ([shift={(3.5cm,-2.5cm)}]current page.center) {\animategraphics[loop,controls,height=2.6cm]{12}{GWs_cross-}{0}{20}};
\end{tikzpicture}
\end{frame}

\begin{frame}{Making gravitational waves}{non-spherical, accelerated motion}
  \alert{Inspirals}, GRBs, bumps on neutron stars, supernovae, the Big Bang. \\%  (BICEP 2).\\
  Gravitational-wave luminosity: \alert{quadrupole} radiation (leading-order)
  \begin{center} $ \boxed{\dot{E}=L_\text{GW} = \f{G}{c^5} \langle (\partial_t^3 \bar{Q}_{ij})^2 \rangle}$\\
  \vspace{2mm}
  \begin{small}$ \quad \bar{Q}_{ij} = Q_{ij}-\frac{1}{3}\delta_{ij}Q^k_k, \quad Q^{ij} = \int \rho(t,{\mathbf x})x^i x^j d^3 x $ \end{small} \end{center}
  \onslide<2->{{\bf Binary systems:} 2 points masses in circular orbit.}
  \begin{tikzpicture}[overlay,remember picture]
    \onslide<5->\node (img0 )[anchor=center,scale=1,opacity=1] at ([shift={(-3.8cm,-2.7cm)}]current page.center) {\animategraphics[loop,controls,height=2.9cm]{12}{Circ_Orb-}{0}{20}};
    \onslide<3->\node (t1 )[anchor=center,scale=1,opacity=1] at ([shift={(2cm,-1.3cm)}]current page.center) {$\mu=m_1 m_2/M^2,\ x(t) = r \cos\Omega t, \ y(t) = r \sin\Omega t $};
    \onslide<4->\node (t2 )[anchor=center,scale=1,opacity=1] at ([shift={(2.25cm,-2.2cm)}]current page.center) {$ Q^{ij} =  2 M\, x^i x^j, \qquad \boxed{\dot{E} = \f{32}{5}\f{G\mu^2}{c^5}r^4 \Omega^6}$};
    \onslide<5->\node (t3 )[anchor=center,scale=1,opacity=1] at ([shift={(2.5cm,-3.2cm)}]current page.center) {Enhancement due to \bl{$e$} (Peters \& Mathews).};
    \onslide<6->\node (t3 )[anchor=center,scale=1,opacity=1] at ([shift={(2cm,-4cm)}]current page.center) {$ \bl{\f{\left(1 + \f{73}{24}e^2+ \f{37}{96} e^4 \right)}{(1-e^2)^{7/2}}}$};
  \end{tikzpicture}
\end{frame}

\begin{frame}{Compact Binary Inspirals}
  \vspace{-3mm}
  Notation:
  \begin{itemize}
    \item $\red{f} $: GW frequency, \hspace{1.5cm} $\Longrightarrow\quad \omega\equiv2\pi f = 2\Omega$.
    %\item $\omega\equiv2\pi f = 2\Omega$: GW angular frequency.
    \item $\red{t}$: observer/detector time \quad$ \Longrightarrow\quad\dot{Y}= dY/dt$.
  \end{itemize}
  \onslide<2->
  {
    Assumptions:
    \begin{itemize}
      \item \bl{Quasi-circular} orbits: $e \ll 1$ and $\tfrac{|\dot{r}|}{r\Omega} <10^{-3} $ at $f=100\,$Hz.
      \item \bl{Quadrupolar} GWs only: higher modes suppressed by $\tfrac{m_1-m_2}{M}\tfrac{v^2}{c^2}$
      \item Use Newtonian/post-Newtonian theory ($\le$3.5{\bf PN}).
      \item \red{Terminate} inspiral at Schwarzschild \red{ISCO} 
      \begin{footnotesize}
      \begin{align*}
      f_\text{ISCO} &= \f{c^3}{6^{3/2}\pi G M} \simeq \bl{1571}\,\text{Hz}\, \left(\f{2.8M_\odot}{M}\right)\hspace{2cm}\\ 
      r_\text{ISCO} &\simeq 25\,\text{km}\left(\frac{M}{2.8 M_\odot}\right)
      \end{align*}
      \end{footnotesize}
    \end{itemize}
   } 
  %
  \onslide<3->
  {
    \vspace{1mm}
    For binary neutron stars ({\bf BNS})
    \begin{itemize}
      \item Point masses: tidal effects for $f \gtrsim 100\,$Hz
      \item Set $m_1 =m_2 = 1.4 M_\odot$, constraints: $1.2 M_\odot \lesssim m_\text{NS}\lesssim 1.6M_\odot$.
      \item Neglect spins: spin-orbit $\tfrac{\dot{E}_\text{SO}}{\dot{E}} \sim \tfrac{v_s}{c}\tfrac{v}{c}\tfrac{R_\text{NS}}{r}$, spin-spin $\sim \tfrac{v^4}{c^4}$
    \end{itemize}
  }
  \onslide<2->
  {
    \begin{tikzpicture}[overlay,remember picture]
    \node (img0 )[anchor=center,scale=1,opacity=1] at ([shift={(4.4cm,-1.5cm)}]current page.center) {\includegraphics[height=3.5cm]{figs/OrbitPotential.pdf}};    
    \end{tikzpicture}
  }  
\end{frame}

@@

\begin{frame}{Compact Binary Inspirals}
\vspace{-3mm}
Integrate $\dot{f} \sim f^{11/3} \quad \Longrightarrow \quad \int dt \sim \int f^{-11/3}df \quad\Rightarrow\quad t \sim \red{f^{-8/3}}$
\\
\vspace{1mm}
Fix integration constant $\tau = t_\text{coal} - t >0$. \\
\vspace{2mm}
Introduce \red{inspiral time}, i.e, time to BNS \bl{coal}escence
%
\begin{small}
\begin{align}
 \red{\tau_\text{insp}(f)} %&= \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,{f^{-8/3}}\nn\\
 &\simeq \bl{16.72\,\text{minutes}} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,{\left(\f{10\,\text{Hz}}{\red{f}}\right)^{\red{8/3}}\quad \red{(2)}}\nn\hspace{2cm}
\end{align}
\end{small}
Number of GW cycles to coalescence: $\int \tfrac{f}{\dot{f}}df\sim \red{f^{-5/3}}$ %$\int df {f}/\dot{f}$
\begin{small}
\begin{align}
 \red{\mathcal{N}_\text{cyc}(f)} %&=\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right]\nn \\
 &\approx 1.605\times \bl{10^4}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}{ \left(\f{10\,\text{Hz}}{\red{f}} \right)^{\red{5/3}}}\quad \red{(3)}\nn\hspace{3.2cm}
\end{align}
\end{small}
%
\begin{tikzpicture}[overlay,remember picture]
  \node[draw, red,thick,minimum width=9.2cm,minimum height=1.3cm] (b) at (5.,3.35){};
  \node[draw, red,thick,minimum width=9cm,minimum height=1.3cm] (b) at (4.4,1){};
  \node (img0 )[anchor=center,scale=1,opacity=1] at ([shift={(3.9cm,-3.2cm)}]current page.center) {\includegraphics[height=3.5cm]{figs/R_of_Tau.pdf}};
\end{tikzpicture}
%
\[ 
\tfrac{\dot{r}}{r}\sim \tfrac{\dot{f}}{f} \Longrightarrow \red{r(\tau)}=r_i \left(\f{\red{\tau}}{\tau_i}\right)^{\red{1/4}} \hspace{4cm}
\]

@@

\begin{frame}{Gravitational waves from BNS}
  TT gauge, multipole, slow-velocity expansions give \todo{add suppl. slide}
  \begin{small}
    \[
    h_{ij}^\text{TT} = \tfrac{2}{D}\tfrac{G}{c^4} \ddot{Q}_{ij}^\text{TT}%(t-z/c)
    \]
    %
    \vspace{-5mm}
    \begin{align*}
    h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi(t)],\\
    h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi(t)],
    \end{align*}
  \end{small}
  %
  %
  $\iota =\,$ orbital \bl{inclination}, and
  \vspace{-3mm}
  %
  \begin{small}
    \begin{align*}
    h_c(t) %&=\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3}\nn\\
    &\approx 1.9\times \bl{10^{-23}} \left(\f{40\,\text{Mpc}}{D}\right)\left(\f{M_c}{1.219 M_\odot}\right)^{5/3}\left(\frac{\red{f}}{10\,\text{Hz}}\right)^{\red{2/3}} \nn\\
    \Phi(t) &=\int_{t_i}^t dt' \omega(t') + \text{PN}
    \end{align*}
  \end{small}\\
  %
  Frequency increases as $f^{11/3}$\\
  Amplitude increases as $f^{10/3} $
  %
  %
  \begin{tikzpicture}[overlay,remember picture]
    \node (img0 )[anchor=center,scale=1,opacity=1] at ([shift={(2.7cm,-3.2cm)}]current page.center) {\includegraphics[height=2.5cm]{figs/Chirp.pdf}};
     \node[anchor=center,scale=1.4,opacity=1] at ([shift={(1cm,-2.4cm)}]current page.center) {\href{https://www.youtube.com/watch?v=CH9JSqUfnSw}{\red{Chirp!}}};
  \end{tikzpicture}
\end{frame}

 \begin{frame}{GW Detectors: \red{Interferometers}}
  Two geodesics separated by $\xi^\mu$:
  %\[
  % \frac{D^2 \xi^\alpha}{D\tau^2} = -R^\alpha_{\beta \gamma\delta}\, \xi^\gamma u^\beta u^\delta 
  %\]
$ \tfrac{d^2 \xi^i}{d\tau^2} = -R^i_{0j0} \xi^j\left(\tfrac{dx^0}{d\tau}\right)^2$ \hfill (detector frame)\\
%\hspace{4cm} $\Longrightarrow \ddot{\xi^i} = -c^2 R^i_{0j0} \xi^j $ \hfill (detector time)\ \\
\vspace{-3mm}
\[
\hspace{4.5cm} \boxed{\ddot{\xi^i} = \frac{1}{2} \ddot{h}_{ij}^\text{TT} \xi^j} \hspace{2.5cm} \text{(detector time)}
\]
%
%
GW along $\mathbf{z}$: 
\[
 h_{ij}^\text{TT} = \left(\begin{array}{cc}
                           h_+ & h_\times \\
                           h_\times & - h_+
                          \end{array}\right) \cos\left(\omega\left( t-\tfrac{z}{c}\right)\right)
\]
Consider mirrors at $(0,0)_1, (L,0)_2$ and $(L,L)_3$
\begin{small}
\begin{align*}
  s_{12} &= \int_1^2 \tfrac{ds}{dx}dx = (x_2-x_1) \left[ 1 + h_+ \cos(\omega t)\right]^{1/2}\hspace{4cm} \\
			      & \approx L \left[ 1+\tfrac{1}{2} h_+ \cos\omega t \right] \\
s_{13} &\approx L \left[ 1-\tfrac{1}{2} h_+ \cos\omega t \right]
\end{align*}
\end{small}
%
\begin{tikzpicture}[overlay,remember picture]
 \node[anchor=center,scale=1,opacity=1] at ([shift={(2cm,-3.4cm)}]current page.center){\todo{Make sure the vid here works}} ;
%\node (img0 )[anchor=center,scale=1,opacity=1] at ([shift={(3.8cm,-2.8cm)}]current page.center) {\animategraphics[loop,controls,height=2.39cm]{12}{LIGO_int-}{0}{102}};
\end{tikzpicture}
 \end{frame}

@@

\begin{frame}{Detection criteria}{Optimal Signal-to-Noise Ratio {\bf SNR}}
\vspace{-5mm}
\begin{footnotesize} 
\[
 \rho=\left[ \int_0^\infty d\ln f\, \f{|\red{2\tilde{h}(f)\sqrt{f}}|^2}{S_n(f)}\right]^{1/2} 
\]
\\
\end{footnotesize}
$\sqrt{S_n(f)}\equiv$ {\it amplitude spectral density} ({\bf ASD}) or  detector noise
\\
\vspace{3mm}
Recall, $\tilde{h}(f) \sqrt{f} \sim f^{-7/6} f^{1/2} = \red{f^{-2/3}}$
\begin{footnotesize} 
\[
 \rho^2 =  (2A)^2 h_0^2 |Q(\theta,\phi,\psi,\iota)|^2 \int_0^{\bl{f_\text{ISCO}}} df \f{f^{-7/3}}{S_n(f)} 
\]
\end{footnotesize}
%
RMS \red{average} over the angles 
\begin{small}
$ \langle \ldots \rangle \equiv \int \sin\theta d\theta d\phi \int d\psi \int\sin\iota d\iota $
\end{small}
\begin{footnotesize} 
\[
 \langle |Q(\theta,\phi,\psi,\iota)|^2 \rangle_\llcorner = \f{4}{25}, \qquad \langle |Q(\theta,\phi,\psi,\iota)|^2 \rangle_\triangle = \f{9}{25}
\]
\end{footnotesize}
%
Sources will have random $\{\theta,\phi,\psi,\iota\}$ \\
\begin{footnotesize} 
\[
0 \le Q(\theta,\phi,\psi,\iota) \le \underbrace{Q(\theta=\tfrac{\pi}{2},\phi,\psi=0,\iota=0)}_{Q_\text{max}}
\]
\end{footnotesize}
% 
\end{frame}
